Gravitational Lens Modelling in a Citizen Science Context

### 2014-10-01 text/ms.tex
edits following comments from AM

The encouraging early results from the first \sw lens search,
carried out on the  $\simeq172$ square degree Canada-France-Hawaii
Telescope Legacy Survey (CFHTLS) imaging by over 30,000 volunteers
(Marshall et al, in prep; More et al, in prep) prompted the question:
could the modelling of the lenses also be done by the volunteers?  If
so, modelling could help assess prioritise lens candidates at an early
stage, which would be very useful with new wide-field and sensitive
surveys, which will yield thousands of lens candidates. There are
several software tools for lens modelling available, and work has
been done on generic interfaces \citep[e.g.][]{2014A&C.....5...28L}.
Indeed, some early designs for \sw included a prototype lens modelling
tool \citep{2010AAS...21543527N}. Moreover, some \sw volunteers are
quite experienced from earlier projects, having spent a thousand hours
or more with data, and are very interested in more demanding
projects.  The interests of citizen-science communities are just
beginning to be studied \citep[e.g.,][]{2013AEdRv..12a0106J}, but it
is clear that some volunteers welcome open-ended challenges, and
sometimes these have led to new scientific results: one example is the
discovery of an exceptional extra-solar planet
\citep{2013ApJ...768..127S}; another is the development of new
algorithms for protein folding \citep{Khatib22112011}.  All these are
grounds for optimism.  There is, however, a basic difficulty in strong
gravitational lensing. Lensed images do not look much like their
source, and still less do they resemble the lensing-mass
distribution.  To model a lens, one needs either to do a lot of random
guessing, or to have a good intuition for what works.

@@

\spl is not part of \sw, but it makes use of the \sw infrastructure,
in particular, the image database and the discussion
forum.\footnote{\url{ http://talk.spacewarps.org}}  The forum is
essential for establishing contact between interested members of the
\sw community and the project science team, and then for enabling
collaboration between them. We were able to collaborate together on
modelling objects from \sw in the usual style of medium-sized
astronomical collaborations, with video-conferencing and in-person
meetings where possible. Preliminary results were immediately
summarized on modelling threads on the forum, and anyone interested
was made welcome to join at any time.
% {\bf Phil asks: how much training was necessary before the volunteers
% understood how to work the interface? Could this training be done with
% no input from the \spl developers, in future? It would be good to
% comment on how scalable this approach to modeling is.}

% PJM: The following paragraph is not accurate: the manipulation of
%  FITS images in the Tools viewer, for example, is done client side.
%  In future I very much hope to see lens modeling enabled in this
%  mode as well!
%
% Another aspect of \sw (and of the Zooniverse family of projects in
% general) that \spl adopts, is to keep all user interaction in a web
% browser, with intensive computing done on a server.  Several
% interesting design and implementation issues arise from these
% considerations, but we will not go into them in this paper,
% concentrating instead on aspects specific to lens modelling.
%

@@

\begin{figure}
  \centering
    \includegraphics[width=0.90\linewidth]{fig/screenshot}
  \caption{Screen grab of \spl in action. A \sw image has been loaded
    in, re-centred and zoomed, and 5 images and associated time delay
    surface ``spaghetti'' contours suggested, using the marking tools
    associated with the buttons along the top of the panel. The mass
    model is optimized server-side when the right-most button is
    pressed.}
  \label{fig:screenshot}
\end{figure}

@@

After examining this feedback, the user can choose to save the
model to the \spl archive, at which point it is assigned an unique
URL.  He can also modify his input and try again, or discard the
attempt altogether.  After archiving, there can be discussion among
modellers, through the \sw forum or by any other means, and revision
of the model.  This is achieved simply by sharing the model's
URL; following its hyperlink takes one to the \spl app, pre-loaded
with the correct data image and input spaghetti. Any archived model
can be revised by any user: they can modify the spaghetti
configuration slightly or drastically, or change options like the size
of the mass tiles. Particularly interesting lens candidates lead to
trees of models in this way.  Discussion among modellers tends to
prune a model tree, focusing attention on the most interesting
models.\footnote{See ``Collaborative gravitational lens
modelling\dots'' in {\tt http://letters.zooniverse.org} for an example.}
% PJM: sadly this Letter is no longer available, Julianne took it
% down. We can post the talk URL instead.
% PS: Letter is available again now.
%\footnote{See the ``collaborative modelling'' thread at
% \url{ http://talk.spacewarps.org} for an example.}

@@

The \sw sims are described in detail in More et al (in prep), but
relevant here is that the sims were of three kinds, as follows.

\begin{enumerate}
  \item Imitating lensed quasars: having a singular elliptical
    isothermal lens (SIE) plus constant external shear, and a circular
    Gaussian source.
  \item Emulating lensed galaxies: similar to the above, but with an
    elliptical de Vaucouleurs source.
  \item Resembling cluster lenses: having a source as above, but a
    more complicated lens, with one dominant elliptical SIE and
    one or more perturbing elliptical SIEs, plus a circular NFW
    \citep{1996ApJ...462..563N,1997ApJ...490..493N} to represent
    the underlying dark matter distribution.
\end{enumerate}
The {\tt gravlens} program \citep{2001astro.ph..2340K} was used.
Formulas for the lenses appear in \cite{2001astro.ph..2341K}. The SIE
lenses follow equations (33--35) of that work, with core radius set to
zero.  The NFW lens is in equations (48) and (50), while shear is the
$\gamma$ term in equation (76).

@@

  \caption[result 6919 (ASW0002z6f)]{Another configuration of arc plus
    counter-image; here the arc is closer to the lensing galaxy than
    the counter-image. (See \secref{example_models} for details.)
    \vspace{1em}}
  \label{fig:6919}
\end{figure}

@@

Each of Figures \ref{fig:6941}--\ref{fig:7022} figures has the
following layout.
$$ \begin{matrix}
\hbox{marked-up image} \qquad &\hbox{model synthetic image} \\
t(x,y)                        &\hbox{model } t(x,y) \\
\kappa(x,y)                   &\hbox{model } \kappa(x,y)
\end{matrix} $$

@@

\Figref{ER_per_sim} shows that \ERf[,rec] of the models tend to be too
high.  However, this is entirely due to the GLASS model density
profiles being too shallow, as illustrated above. We can separate out
the performance of the \spl interface and its users by comparing their
results with the  Einstein radii of \spl models made by an expert
(PS). Discounting the models which were flagged by the volunteers as
poor, the mean Einstein radius overestimate was 10\%, with a 15\%
standard deviation (shown by the light grey band in \Figref{ER_per_sim}).
The expert models show a similar bias,
with standard deviation 10\% (the dark grey band in \Figref{ER_per_sim}).
One source of this systematic error is that it is difficult to center
of the lens accurately: an offset leads to a flatter mass profile for
the model compared to the simulation.

@@

There is, however, plenty of room for improvement:
\begin{enumerate}
\item \spl tends to overestimate the Einstein radius (evident from
  \Figref{ER_per_sim}), and it is also apparent that the model density
  profiles tend to be too shallow.  The likely explanation is that
  while the sims are steeply peaked at the centre, the pixellated mass
  model fixes a comparatively large area near the central at constant
  density. Allowing smaller pixels in the centre region, thus enabling
  a steeper centre \citep[similar to the ``high resolution'' feature
    implemented in][]{2014arXiv1401.7990C} may remove this bias.  The
  use of simply parameterised, appropriately steeply-profiled models
  would also avoid the problem.
\item Currently, \spl does not attempt to model the source shape; the
  user identifies the brightest points on the image, and these are
  taken as images of a point-like source, whose positions must be
  reproduced exactly. For generating a synthetic image, a conical
  source profile is assumed. Fitting for the source profile to
  optimize resemblance to the observed lensed image after the lens
  model has been generated, is algorithmically straightforward
  \citep[cf.][]{2003ApJ...590..673W,2006MNRAS.371..983S} and planned
  to be implemented.  This would alleviate another problem with \spl,
  which is that there is as yet no quantitative figure of merit for
  any given model: assessment of each model is a judgment call based
  on the synthetic image, and on whether the mass distribution and the
  arrival-time surface show suspicious features.
\item Another limitation so far in \spl is that the lens is assumed to
  be dominated by one galaxy, which puts most galaxy-group lenses
  beyond the reach of the modeler. Since complicated group lenses are
  some of the most interesting candidates present, removing this
  limitation is most desirable.  From the users' point of view, it
  would mean that spaghetti contours with more than one maximum can be
  allowed.  For examples, see Figure 5c in \citep{2001ApJ...557..594R}
  and Figure 4b in \cite{2003ApJ...590...39K}.
\item At present, a single false-color composite is used as the data.
  An option could be added to use all available filters, individually
  or in combination, at the user wishes.
\item As mentioned above, the option of revising an already-archived
  model is already available.  Desired now are tools for comparing
  different models of a given system, both visually and through
  different statistical measures.  As evidenced by a current
  collaborative modelling effort, a particularly interesting candidate
  can lead to an extended discussion and dozens of models, that
  in some way sample the high likelihood region of model parameter
  space.
\item Better tutorial materials are also needed, and this would
  address some of the problem areas found in the modelling challenge.
  For example, we saw in \secref{tests.t1} that volunteers are most
  prone to making errors in two situations: when in identifying an arc-
  like structure while placing the points, and in identifying the
  correct ordering of the points in nearly-symmetric configurations.
  Better and more detailed introductory materials would also allow
  the community of modellers to grow faster and without individual
  instructions by experts or experienced volunteers.
\end{enumerate}

The \spl program was developed by Küng, with design suggestions from
Coles, Cornen and Saha, and feedback from all co-authors.  The
simulations were created by A.~More, in consultation with Marshall,
S.~More and Verma.  Modelling was done by Baeten, Cornen, Macmillan,
Odermatt, Saha and Wilcox, with post-modelling analysis by Kueng and
Saha.  All authors participated in writing and editing the manuscript.

@@

We thank the Swiss Society for Astrophysics and Astronomy and the
Swiss Academy of Sciences. AV is supported by a research fellowship
from the Leverhulme Trust. This work was supported in part by the U.S.
Department of Energy under contract number DE-AC02-76SF00515.

>>>>

The encouraging early results from the first \sw lens search, carried
out on the $\simeq172$ square degree Canada-France-Hawaii Telescope
Legacy Survey (CFHTLS) imaging by over 30,000 volunteers (Marshall et
al, in prep; More et al, in prep) prompted the question: could the
modelling of the lenses also be done by the volunteers?  If so,
modelling could help assess prioritise lens candidates at an early
stage, which would be very useful with new wide-field and sensitive
surveys, which will yield thousands of lens candidates.  There are
several software tools for lens modelling available, and work has been
done on generic interfaces \citep[e.g.][]{2014A&C.....5...28L}.  Some
early designs for \sw included a prototype lens modelling tool
\citep{2010AAS...21543527N}. Moreover, some \sw volunteers are quite
experienced from earlier projects, having spent a thousand hours or
more with data, and are very interested in more demanding projects.
The interests of citizen-science communities are just beginning to be
studied \citep[e.g.,][]{2013AEdRv..12a0106J}, but it is clear that
some volunteers welcome open-ended challenges, and sometimes these
have led to new scientific results: one example is the discovery of an
exceptional extra-solar planet \citep{2013ApJ...768..127S}; another is
the development of new algorithms for protein folding
\citep{Khatib22112011}.  All these are grounds for optimism.  There
is, however, a basic difficulty in strong gravitational
lensing. Lensed images do not look much like their source, and still
less do they resemble the lensing-mass distribution.  To model a lens,
one needs either to do a lot of random guessing, or to have a good
intuition for what works.

@@

\spl is a mass modelling program that makes use of the \sw
infrastructure, in particular, the image database and the discussion
forum.\footnote{\url{ http://talk.spacewarps.org}} The forum is
essential for establishing contact between interested members of the
\sw community and the project science team, and then for enabling
collaboration between them. We were able to collaborate together on
modelling objects from \sw in the usual style of medium-sized
astronomical collaborations, with video-conferencing and in-person
meetings where possible. Preliminary results were immediately
summarized on modelling threads on the forum, and anyone interested
was made welcome to join at any time.

@@

\begin{figure}
  \centering
    \includegraphics[width=0.90\linewidth]{fig/screenshot}
  \caption{Screen grab of \spl in action. A \sw image has been loaded
    in, re-centred and zoomed.  Five images and the associated
    ``spaghetti'' contours have then been suggested, using the marking
    tools associated with the buttons along the top of the panel. The
    mass model is generated server-side when the right-most button is
    pressed.}
  \label{fig:screenshot}
\end{figure}

@@

After examining this feedback, the user can choose to save the
model to the \spl archive, at which point it is assigned an unique
URL.  He can also modify his input and try again, or discard the
attempt altogether.  After archiving, there can be discussion among
modellers, through the \sw forum or by any other means, and revision
of the model.  This is achieved simply by sharing the model's
URL; following its hyperlink takes one to the \spl app, pre-loaded
with the correct data image and input spaghetti. Any archived model
can be revised by any user: they can modify the spaghetti
configuration slightly or drastically, or change options like the size
of the mass tiles. Particularly interesting lens candidates lead to
trees of models in this way.  Discussion among modellers tends to
prune a model tree, focusing attention on the most interesting
models.\footnote{See ``Collaborative gravitational lens
modelling\dots'' in {\tt http://letters.zooniverse.org} for an example.}
% This Letter was inaccessible for a while is available again now.

@@

The \sw sims are described in detail in More et al (in prep), but
relevant here is that the sims were of three kinds, as follows.
\begin{enumerate}
\item Lensed quasars: The lens is modelled as a singular isothermal
  ellipsoid (SIE) and a constant external shear whereas the quasar is
  represented with a circular Gaussian source whose size is given by
  the point spread function (PSF) in each imaging band.
\item Galaxy-scale lenses: The lens model is the same as above whereas
  the background galaxy is modelled as an elliptical de~Vaucouleurs.
\item Groups-scale lenses: The lens model includes SIE models for the
  central galaxy and the inner group members, plus a circular NFW
  (ref) to represent the underlying dark matter distribution and the
  background galaxy model stays the same as galaxy-scale lenses.
\end{enumerate}
The {\tt gravlens} program \citep{2001astro.ph..2340K} was used.
Formulas for the lenses appear in \cite{2001astro.ph..2341K}. The SIE
lenses follow equations (33--35) of that work, with core radius set to
zero.  The NFW lens is in equations (48) and (50), while shear is the
$\gamma$ term in equation (76).

@@

  \caption[result 6919 (ASW0002z6f)]{Another configuration of arc plus
    counter-image: an arc and counter image where the arc is closer to
    the lensing galaxy than the counter image. (See
    \secref{example_models} for details.)
    \vspace{1em}}
  \label{fig:6919}
\end{figure}

@@

Each of Figures \ref{fig:6941}--\ref{fig:7022} figures has the
following layout.
$$ \begin{matrix}
\hbox{marked-up CFHTLS image} \qquad &\hbox{model synthetic image} \\
t(x,y)                               &\hbox{model } t(x,y) \\
\kappa(x,y)                          &\hbox{model } \kappa(x,y)
\end{matrix} $$

@@

\Figref{ER_per_sim} shows that \ERf[,rec] of the models tend to be too
high.  However, this is entirely due to the GLASS model density
profiles being too shallow, as illustrated above. We can separate out
the performance of the \spl interface and its users by comparing their
results with the Einstein radii of \spl models made by an expert
(PS). Discounting the models which were flagged by the volunteers as
poor, the mean Einstein radius overestimate was 10\%, with a 15\%
standard deviation (shown by the light grey band in
\Figref{ER_per_sim}).  The expert models show a similar bias, with
standard deviation 10\% (the dark grey band in \Figref{ER_per_sim}).
One source of this systematic error is that it is difficult to centre
the lens accurately: an offset leads to a flatter mass profile for the
model compared to the simulation.

@@

There is, however, plenty of room for improvement:
\begin{enumerate}
\item \spl tends to overestimate the Einstein radius (evident from
  \Figref{ER_per_sim}), and the model density profiles tend to be
  shallower than the SIE model used for generating the sims.  The
  likely explanation is that while the sims are steeply peaked at the
  centre, the pixellated mass model fixes a comparatively large area
  near the central at constant density. Allowing smaller pixels in the
  centre region, thus enabling a steeper centre \citep[similar to the
    ``high resolution'' feature implemented in][]{2014arXiv1401.7990C}
  may remove this bias.  The use of simply parameterised,
  appropriately steeply-profiled models would also avoid the problem.
\item Currently, \spl does not attempt to model the source shape; the
  user identifies the brightest points on the image, and these are
  taken as images of a point-like source, whose positions must be
  reproduced exactly. For generating a synthetic image, a conical
  source profile is assumed. Fitting for the source profile to
  optimize resemblance to the observed lensed image after the lens
  model has been generated, is algorithmically straightforward
  \citep[cf.][]{2003ApJ...590..673W,2006MNRAS.371..983S} and planned
  to be implemented.  This would alleviate another problem with \spl,
  which is that there is as yet no quantitative figure of merit for
  any given model: assessment of each model is a judgment call based
  on the synthetic image, and on whether the mass distribution and the
  arrival-time surface show suspicious features.  Another possibility
  would be use the \spl models as a feeder to a different
  lens-modelling program which already implements source-profile
  fitting.
\item Another limitation so far in \spl is that the lens is assumed to
  be dominated by one galaxy, which puts most galaxy-group lenses
  beyond the reach of the modeler. Since complicated group lenses are
  some of the most interesting candidates present, removing this
  limitation is most desirable.  From the users' point of view, it
  would mean that spaghetti contours with more than one maximum can be
  allowed.  For examples, see Figure 5c in \citep{2001ApJ...557..594R}
  and Figure 4b in \cite{2003ApJ...590...39K}.
\item At present, a single false-color composite is used as the data.
  An option could be added to use all available filters, individually
  or in combination, at the user wishes.
\item As mentioned above, the option of revising an already-archived
  model is already available.  Desired now are tools for comparing
  different models of a given system, both visually and through
  different statistical measures.  As evidenced by a current
  collaborative modelling effort, a particularly interesting candidate
  can lead to an extended discussion and dozens of models, that
  in some way sample the high likelihood region of model parameter
  space.
\item Better tutorial materials are also needed, and this would
  address some of the problem areas found in the modelling challenge.
  For example, we saw in \secref{tests.t1} that volunteers are most
  prone to making errors in two situations: when in identifying an arc-
  like structure while placing the points, and in identifying the
  correct ordering of the points in nearly-symmetric configurations.
  Better and more detailed introductory materials would also allow
  the community of modellers to grow faster and without individual
  instructions by experts or experienced volunteers.
\end{enumerate}

The \spl program was developed by Küng, with design suggestions from
Coles, Cornen and Saha, and feedback from all co-authors.  The
simulations were created by A.~More, in consultation with Marshall,
S.~More and Verma.  Modelling was done by Baeten, Cornen, Macmillan,
Odermatt, Saha and Wilcox, with post-modelling analysis by Küng and
Saha.  All authors participated in writing and editing the manuscript.

@@

We thank the Swiss Society for Astrophysics and Astronomy and the
Swiss Academy of Sciences. The work of AM and SM was supported by
World Premier International Research Center Initiative (WPI
Initiative), MEXT, Japan. AV is supported by a research fellowship
from the Leverhulme Trust. This work was supported in part by the U.S.
Department of Energy under contract number DE-AC02-76SF00515.
